---
title: "Spatial@iDiv - Day 3"
author: "Isabel Rosa"
date: "07/11/2018"
output: html_document
---


## Operations with both Rasters and Shapefiles
Here I'll show a couple of examples of operation that use feature data as inputs and output rasters:
Distance to features - calculates the euclidean distance from each cell/pixel to the closest feature (e.g., roads, settlements). Outputs a raster file with these distances.
Interpolation: a world in itself! Very vey short example provided here (based on a single method, IDW), please see more here: http://www.rspatial.org/analysis/rst/4-interpolation.html
To better understand interpolation I advise you to read first about spatial autocorrelation: http://www.rspatial.org/analysis/rst/3-spauto.html

To use interpolation metrics you need to load another packaged called gstat
Inverse distance weighted (IDW) - See more also here: http://desktop.arcgis.com/en/arcmap/10.3/tools/3d-analyst-toolbox/how-idw-works.htm

```{r echo=TRUE}

# create an empty raster (little trick using existing raster)
dist_sett<-tc*0
# or you can create an empty one like before
# dist_sett <- raster(ncol=ncol(tc), nrow=nrow(tc), xmx=extent(tc)@xmax, xmn=extent(tc)@xmin, ymn=extent(tc)@ymin, ymx=extent(tc)@ymax)

# Distance to points
dist_sett <- distanceFromPoints(dist_sett, sc_sett)

# you can then mask the outside area of Santa Catarina
dist_sett <- mask(dist_sett, tc)

# plot results
plot(dist_sett, main = "Distance to settlements (m)")

# load gstat
library(gstat)
idw_sett<-tc*0

# compute the model, see reference for more detail
gs <- gstat(formula=population~1, locations=sc_sett, nmax=5, set=list(idp = 2))
idw_out <- interpolate(idw_sett, gs)

## [inverse distance weighted interpolation]
sc_pop <- mask(idw_out, tc)
plot(sc_pop, main = "Santa Catarina Population")

```


## Export Shapefiles and Rasters
It's very easy to export both shapefiles and rasters from R to be visualized in QGIS or ArcMap. 

```{r echo=TRUE}

# Save feature layers (point, polygon, polyline) to shapefile 
writeOGR(sc_largesett, dsn=".", layer="SC_largeSett", driver="ESRI Shapefile" )

#Exporting raster
writeRaster(sc_pop, filename="SC_popmap", format="GTiff" )

```






Advanced Vector Data Operations I (ArcMap)
 - find spatial vector data
 - join tables (countries, statistics, species ranges)
 - pick elevation data from raster to points
 - basic statistical analyses (t-test with raster distribution and point data distribution)
 - spatial autocorrelation + semivariogram
coffee break
Same as above but in R
  - explore analyses a bit further
 - plot some stuff with ggplot


Lunch



Introduction to interpolation methods (ArcMap and R)
  - IDW (using data prepared in the morning)
(This session will only be in R)

 - Krigging (using data prepared in the morning)
(This session will only be in R)


